Overlooked Dynamics in Algorithmic Recourse

### 2022-07-08 paper/sections/methodology.qmd
moving to julia

In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro. Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks. We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
$$
\mathbf{s}^\prime = \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  \ell(M(f(\mathbf{s}^\prime)),t)^\mathsf{T} \mathbf{1}_K + \lambda h(f(\mathbf{s}^\prime)) \right\}
$$ {#eq-general}

Here $\mathbf{s}^\prime$ is a $K$-dimensional vector of counterfactual states and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one. This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared. In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].

@@

A noteworthy practical consideration is the choice of $T$ and $B$. The higher these values, the more factual instances undergo recourse throughout the entire experiment. Of course, this is likely to lead to more pronounced domain and model shifts by time $T$. At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm’s decisions. In our experiments, we choose the values such that $T \cdot B$ corresponds to the application of recourse on $25-50\%$ of the negative instances from the initial dataset. As we collect data at each time $t$, we can also verify the impact of recourse when it is implemented for a smaller number of individuals. Using our framework extenson of @pawelczyk2021carla the experiment can be conducted on an arbitrary number of algorithmic recourse generators in parallel. As all generators make use of the same initial model and initial dataset, the differences in domain and model shifts observed throughout the rounds depend solely on the employed generator.

>>>>

In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro. Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks. We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:

[AGREE ON NOTATION]

$$
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left(  {\left\{\ell(M(f(s_k^\prime)),t)\right\}_K}^\mathsf{T} \mathbf{1}_K + \lambda {\left\{h(f(s_k^\prime)) \right\}_K}^\mathsf{T} \mathbf{1}_K \right) \\
&= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \left(\ell . ((M\circ f).(\mathbf{s}^\prime),t) + \lambda (h \circ f).(\mathbf{s}^\prime)\right)^\mathsf{T} \mathbf{1}_K \right\} \\
&= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)}+ \lambda {h(f(s_k^\prime)) }  \right\}
\end{aligned}
$$ {#eq-general}

Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one. This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared. In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].

@@

A noteworthy practical consideration is the choice of $T$ and $B$. The higher these values, the more factual instances undergo recourse throughout the entire experiment. Of course, this is likely to lead to more pronounced domain and model shifts by time $T$. At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm’s decisions. In our experiments, we choose the values such that $T \cdot B$ corresponds to the application of recourse on $25-50\%$ of the negative instances from the initial dataset. As we collect data at each time $t$, we can also verify the impact of recourse when it is implemented for a smaller number of individuals. Using our framework the experiment can be conducted on an arbitrary number of algorithmic recourse generators. As all generators make use of the same initial model and initial dataset, the differences in domain and model shifts observed throughout the rounds depend solely on the employed generator.

### 2022-07-08 paper/sections/related.qmd
moving to julia

...

>>>>

Despite the large and growing number of approaches to counterfactual search there have been surprisingly few benchmark studies that compare different methodologies. This may be partially due to limited software availability in this space. Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework that can be used to apply and benchmark different methodology; finally, @altmeyer2022CounterfactualExplanations provides an extensible and language-agnostic implementation in Julia. For the experiments conducted in this work we relied on the latter.
